agdarsec -- Total Parser Combinators

### 2017-04-18 doc/tyde17/main.tex
[ lhs2tex ] capturing code and including it
[ agda ] capturing code and including it
Had to modify some of the source files to:
- hide implicit Set arguments using sections
- add tags to use capture the code

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

@@

@@

@@

% insert anyChar

@@

% insert guard, digit

@@

% insert guardM, digit

@@

We have seen how we can already rely on the standard library
of the host language to seemlessly implement combinators.
We can leverage even more of the existing codebase by noticing
that the type constructor \type{Parser} is a \type{Functor},
an \type{Applicative}, a \type{Monad} and also an \type{Alternative}.

@@

% insert some, many

@@

The problem already shows up in the definition of \parser{some}
which will only make progress if its argument actually uses up
part of the input string. Otherwise it may loop. However this
is not the typical hurdle programmers stumble upon: demanding
a non empty list of nothing at all is after all rather silly.
The issue manifests itself naturally whenever defining a left
recursive grammar which leads us to introducing the prototypical
such example \type{Expr}:

@@

If we simply reproduce this definition,

@@

>>>>

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{agda}
\usepackage{todonotes}
\usepackage{catchfilebetweentags}

\newcommand{\shgrab}[1]{\bigskip\ExecuteMetaData[parser.tex]{#1}}
\newcommand{\hgrab}[1]{\ExecuteMetaData[parser.tex]{#1}}

\newcommand{\sagrab}[2]{\bigskip\ExecuteMetaData[#1.tex]{#2}}
\newcommand{\agrab}[2]{\ExecuteMetaData[#1.tex]{#2}}

\input{commands.tex}
\input{lhs2tex-prelude.tex}

@@

\shgrab{parser}

@@

\shgrab{parse}

@@

\shgrab{anychar}

@@

\shgrab{guard}
\hgrab{digit}

@@

\shgrab{guard2}
\hgrab{digit2}

@@

We have seen how we can already rely on the standard library
of the host language to seemlessly implement combinators.
We can leverage even more of the existing codebase by noticing
that the type constructor \type{Parser} is a \type{Functor},
an \type{Applicative}~\todo{cite}, a \type{Monad} and also an \type{Alternative}.

@@

\shgrab{some}
\hgrab{many}

@@

The problem already shows up in the definition of \parser{some}
which will only make progress if its argument actually uses up
part of the input string. Otherwise it may loop. However this
is not the typical hurdle programmers stumble upon: demanding
a non empty list of nothing at all is after all rather silly.
The issue manifests itself naturally whenever defining a left
recursive grammar which leads us to introducing the prototypical
such example: \type{Expr}, a minimal language of arithmetic
expressions.

@@

The intuitive solution is to simply reproduce this definition by
introducing an inductive type for \type{Expr} and then defining
the parser as an alternative between a literal on one hand and a
sub-expression, the character '+', and another sub-expression on
the other.

\shgrab{Expr}
\hgrab{expr}

However this leads to an infinite loop. Indeed, the second
alternative performs a recursive call to \parser{expr} even
though it hasn't consumed any character from the input string.

The typical solution to this problem is to introduce two 'tiers'
of expressions: the \parser{base} ones which can only be whole
expressions if we consume an opening parenthesis first and the
\parser{expr'} ones which are left-associated chains of \parser{base}
expressions connected by '+'.

\shgrab{base}
\hgrab{expr2}

This approach can be generalised when defining more complex
languages by having even more tiers, one for each
\emph{fixity level}. An extended language of arithmetic
expressions would for instance distinguish the level at
which addition and substraction live from the one at which
multiplication and division do.

Our issue with this solution is twofold. Firt, although we did
eventually managed to build a parser that worked as expected,
the compiler was unable to warn us and guide us towards this
correct solution. Additionnally, the blatant partiality of
some of these definitions means that these combinators and
these types are wholly unsuitable in a total. We could, of
course use an escape hatch and implement our parsers in Haskell
but that would both be unsafe and mean we would not be able
to run them at typechecking time which we may want to do if
we embed checked examples in our software's documentation,
or use constant values (e.g. filepaths).

@@

Our implementation of Total Parser Combinators is in Agda,
a total dependently typed programming language and it will
rely heavily on indexed sets. But the indices will not be
playing any interesting role apart from enforcing totality.
As a consequence, we introduce combinators to build indexed
sets without having to mention the index explicitly. This
ought to make the types more readable by focusing on the
important components and hiding away the artefacts of the
encoding.

The first kind of combinator corresponds to operations on
sets which are lifted to indexed sets by silently propagating
the index. It may be a bit surprising to see infix notations
(in Agda underscores correspond to positions in which arguments
are to be inserted) for functions taking three arguments but
they are only meant to be partially applied.

\sagrab{Relation/Unary/Indexed}{arrow}
\agrab{Relation/Unary/Indexed}{product}

The second kind turns an indexed set into a set by universally
quantifying over the index. Here we use curly braces so that
the index we use is an \emph{implicit} argument we will never
have to write.

\sagrab{Relation/Unary/Indexed}{forall}

Finally, the last kind of combinator\todo{cite}
Instead of defining it as a

\sagrab{Induction/Nat/Strong}{box}

\sagrab{Induction/Nat/Strong}{extract}
\agrab{Induction/Nat/Strong}{duplicate}

\sagrab{Induction/Nat/Strong}{map}
\agrab{Induction/Nat/Strong}{app}

\agrab{Induction/Nat/Strong}{fix}

\paragraph{Remark} The type of \AF{fix} corresponds exactly to
strong induction for the natural numbers. A similar \AR{□} type
constructor can be defined for any induction principle relying
on an accessibility predicate. Which means that a library's
types can be cleaned up by using these combinators in any
situation where one had to give up structural induction for a
more powerful alternative.

\section{Parsing, Totally}
